---
title: "Model Fitting and Ancestral Character State Reconstruction"
output:
  html_document:
    df_print: paged
---

**Aim**: Test if a simulated character evolves under symmetric or asymmetric Markov model of character evolution using AIC. Reconstruct its ancestral states.

1. Simulate a tree
2. Simulate character evolution under asymmetric model
3. Fit symmetric and asymmetric models
4. Reconstruct character's ancestral states
5. Compare models using AIC


Load necessary packages
```{r}
# uncomment if you do not have this packages installe
#install.packages("corHMM")
#install.packages("phytools")

library("corHMM")
library("phytools")

```

## Simulate a Tree and Traits
```{r}
# simulate tree using pure birth process
tree<-pbtree(n=200, scale=100, b=1, d=0)
plot(tree)

# simulate character evolution
# make rate matrix Q
Q <- matrix(
  c(
    -0.03, 0.03,
    0.1, -0.1
    ), 2,2, byrow = T)

Q

```

Simulate character evolution on tree using Q
```{r}
hist <- sim.history(tree, Q, nsim=1)
plot(hist)
```


## Inference using Symmetrical Model

Now construct rate matrices that show parametrization schemes for ML inference.
```{r}
# one-rate symmetrical model

Q.Sym <- matrix(
  c(
    NA, 1,
    1, NA
  ), 2,2, byrow = T)

Q.Sym

# Inference
taxa <- cbind(hist$tip.label, hist$states)
Recon_Q.Sim <- rayDISC(hist, taxa, rate.mat=Q.Sym, node.states="marginal", 
        model="ARD", root.p="maddfitz")

# infered rate matrix
Recon_Q.Sim

# plot ANCE
plotRECON(tree, Recon_Q.Sim$states, piecolors=c('black', 'red'), title="1-rate Model")


```

## Inference using Asymmetrical Model   
```{r}
# 2-rate symmetrical model

Q.Asym <- matrix(
  c(
    NA, 1,
    2, NA
  ), 2,2, byrow = T)

Q.Asym

# Inference
taxa <- cbind(hist$tip.label, hist$states)
Recon_Q.Asim <- rayDISC(hist, taxa, rate.mat=Q.Asym, node.states="marginal", 
                       model="ARD", root.p="maddfitz")

# infered rate matrix
Recon_Q.Asim

# plot ANCE
plotRECON(tree, Recon_Q.Asim$states, piecolors=c('black', 'red'), title="1-rate Model")


```

## Compare Results 

Compare Model Fit using AIC 
```{r}
Q
Recon_Q.Sim
Recon_Q.Asim

# delta AIC
Recon_Q.Sim$AIC-Recon_Q.Asim$AIC
```


## Home Excrcise
Try to code your own Likelihood function that calculates likelihood of a state pattern given (i) a tree, (ii) rate matrix with specified parameters, and (iii) a specified initial vector. It can be a simple Ln function that works specifically with small trees (3 taxa). Your toy algorithm should work like Felsentein's pruning algorithm from Lecture 6. You may choose any states at tips (use binary characters for brevity).

Before, coding Ln function, simulate a tree and note the main data structures that store tree topology and branch lengths

```{r}
tr<-pbtree(n=3, scale=1, b=1, d=0)
plot(tr)
nodelabels()
edgelabels()
tiplabels()
```
You will need thee object for you Ln function *tr$edge* *tr$edge.length*
```{r}
# tree topologu
tr$edge

# branch lengths
tr$edge.length

```

