---
title: "Model Fitting and Ancestral Character State Reconstruction"
output:
  html_document:
    df_print: paged
---

**Aim**: Test if a simulated character evolves under symmetric or asymmetric Markov model of character evolution using AIC. Reconstruct its ancestral states.

1. Simulate a tree
2. Simulate character evolution under asymmetric model
3. Fit symmetric and asymmetric models
4. Reconstruct character's ancestral states
5. Compare models using AIC


Load necessary packages
```{r}
# uncomment if you do not have this packages installe
#install.packages("corHMM")
#install.packages("phytools")

library("corHMM")
library("phytools")

```

## Simulate a Tree and Traits
```{r}
# simulate tree using pure birth process
tree<-pbtree(n=200, scale=100, b=1, d=0)
plot(tree)

# simulate character evolution
# make rate matrix Q
Q <- matrix(
  c(
    -0.03, 0.03,
    0.1, -0.1
    ), 2,2, byrow = T)

Q

```

Simulate character evolution on tree using Q
```{r}
hist <- sim.history(tree, Q, nsim=1)
plot(hist)
```


## Inference using Symmetrical Model

Now construct rate matrices that show parametrization schemes for ML inference.
```{r}
# one-rate symmetrical model

Q.Sym <- matrix(
  c(
    NA, 1,
    1, NA
  ), 2,2, byrow = T)

Q.Sym

# Inference
taxa <- cbind(hist$tip.label, hist$states)
Recon_Q.Sim <- rayDISC(hist, taxa, rate.mat=Q.Sym, node.states="marginal", 
        model="ARD", root.p="maddfitz")

# infered rate matrix
Recon_Q.Sim

# plot ANCE
plotRECON(tree, Recon_Q.Sim$states, piecolors=c('black', 'red'), title="1-rate Model")


```

## Inference using Asymmetrical Model   
```{r}
# 2-rate symmetrical model

Q.Asym <- matrix(
  c(
    NA, 1,
    2, NA
  ), 2,2, byrow = T)

Q.Asym

# Inference
taxa <- cbind(hist$tip.label, hist$states)
Recon_Q.Asim <- rayDISC(hist, taxa, rate.mat=Q.Asym, node.states="marginal", 
                       model="ARD", root.p="maddfitz")

# infered rate matrix
Recon_Q.Asim

# plot ANCE
plotRECON(tree, Recon_Q.Asim$states, piecolors=c('black', 'red'), title="1-rate Model")


```

## Compare Results 

Compare Model Fit using AIC 
```{r}
Q
Recon_Q.Sim
Recon_Q.Asim

# delta AIC
Recon_Q.Sim$AIC-Recon_Q.Asim$AIC
```


## Home Excrcise
Try to code your own Likelihood function that calculates likelihood of a state pattern given (i) a tree, (ii) rate matrix with specified parameters, and (iii) a specified initial vector. It can be a simple Ln function that works specifically with small trees (3 taxa). Your toy algorithm should work like Felsentein's pruning algorithm from Lecture 6. You may choose any states at tips (use binary characters for brevity).

Before, coding Ln function, simulate a tree and note the main data structures that store tree topology and branch lengths

```{r}
set.seed(100)
tr<-pbtree(n=3, scale=1, b=1, d=0)
plot(tr)
nodelabels()
edgelabels()
tiplabels()

# save tree file
library(ape)
write.tree(tr, file='3_taxon.tree')
```
You will need thee object for you Ln function *tr$edge* *tr$edge.length*
```{r}
# tree topologu
tr$edge

# branch lengths
tr$edge.length

```

### Calculating Likelihood on tree
See lecture 6 for Felsentein's pruning algorithm.

The general steps of the algorithm are as follows:

1. Initialize $Q$ matrix and $\pi$ (initial vector)
2. Recode data at tips as probabilities
3. Calculate conditional likelihoods $L_c$ at each node by traversing the tree from tips to roots. For two branches -- left and right ($b_l$ and $b_r$) -- and their corresponding tip probabilities ($T_l$, $T_r$, both column vectors), the conditional likelihood is  $L_c=e^{Qb_l}T_l*e^{Qb_r}T_r$. Since R does not have inbuilt functionality for column vectors, $T_l$, $T_r$ can be treated as row vectors through the following conversion $L_c=T_l(e^{Qb_l})^\textbf{T}*T_r(e^{Qb_r})^\textbf{T}$, where $^\textbf{T}$ is a matrix transpose
4. Upon reaching root, the tree likelihood is $L=sum(L_c\pi)$


#### 1. Initializing Q and pi
```{r}

# install packages exmp if you do not have it
library (expm)

Q <- matrix(
  c(
    -0.1, 0.1,
    0.1, -0.1
    ), 2,2, byrow = T)

pi=c(0.5, 0.5)
Q
pi

```



#### 2. Recoding data at tips as probabilities
```{r}
# read in tree
tr <- read.tree(file='3_taxon.tree')

# Let's imagine that our taxa at tips (t1, t2, t3) have the following character states (we use a binary character)
data <- c(1,0,1)
names(data) <- c('t1', 't2', 't3')
data
```

We rate their tip probability vectors as matrix
```{r}
data.probs <- matrix(c(
  0,1,
  1,0,
  0,1
  ), 3,2,byrow=T
)
rownames(data.probs) <- c('t1', 't2', 't3')
data.probs
```

#### 3. Calculating conditional likelihoods
```{r}
# Let's plot a tree to see its structure that is encoded on the objects tr$edge and tr$edge.length
plot(tr)
nodelabels()
edgelabels()
tiplabels()
```


First we calculate Lc for node 5 with tips t3 and t2
```{r}
# get ids of two branches node5 -> t3 and node5 -> t2 in tr$edge object (they are 3 and 2)
t3t2.ids=which(tr$edge[,1]==5)

# get branch lengths for those branches
br_l_r=tr$edge.length[t3t2.ids]

# calculate conditional likelihood using matrix exponential expm(), and matrix multiplication %*%
L5=data.probs[3,] %*% t(expm(Q*br_l_r[1]))*(data.probs[2,]%*%t(expm(Q*br_l_r[2])))
L5
```

Now we calculate root likelihood (node4)
```{r}
# get ids of two branches node4 -> t1 and node4 -> node5 
n5t1.ids=which(tr$edge[,1]==4)
# get branch lengths for those branches
br_l_r=tr$edge.length[n5t1.ids]
br_l_r

# calculate conditional likelihood at root using matrix exponential expm(), and matrix multiplication %*%
Lr=(data.probs[1,]%*%t(expm(Q*br_l_r[1])))*(L5 %*% t(expm(Q*br_l_r[2])))
Lr
```

#### 4. Calculating tree likelihood
```{r}
L=sum(Lr*pi)

# likelihood and Log likelihood of the tree
L
log(L)
```

##  Potential Project for this course !!!

Coding a general algorithm that can calculate Ln of any tree and perform maximum likelihood inference (using something like *optimize()* function as in the binomial tutorial) can be your project for this course!

